### 2020-03-02 doc/icfp20/intro.tex
tweak intro

Edgar Var\`{e}se describes music as ``organized sound'', and
throughout history cultures have created and applied systems of rules
and guidelines to govern the music they create, most notably the
Common Practice Period of Western music spanning the 17th to early
20th centuries. These rule systems are seldom absolute, and indeed
deliberate breaking of the rules is often part of the aesthetic, but they
roughly constrain the music they apply to and
give it a common form and sound.

@@

To describe rules of basic harmonic structure,
\citet{magalhaes-harmtrace} and its successors use dependent types,
for example to index chords by major or minor mode.  However Haskell
currently has limited support for dependent types, and requires many
extentions and tricks such as the use of singleton
types~\citep{eisenberg-singleton}.

In this paper we explore what can be done in the context of music by
using a programming language that offers full dependent types. We use
the language Agda~\citep{norell-phd} since it is fairly mature and
aims to be a functional programming language as well as a proof
assistant. It also features a Haskell FFI so we can take advantage of
existing Haskell libraries, in partiular for MIDI.

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
formalize it in type theory. However this is difficult since most
``rules'' in music theory are not absolute, but rather suggestions of
varying degrees (themselves vaguely specified) of importance. Since
ultimately the resulting created music is more important than the
theory, an alternative is to attempt to indirectly deduce the rules
from music samples, using for example machine
learning~\citep{huang-cp}. However there is the danger that since
the rules were never explicit they could end up being violated with
poor results; for example parallel fifths and octaves that are
extremely rare (and almost always due to special allowed
circumstances) in Bach chorales used as training data resulted
in a much higher frequency of their use in generated
harmonizations~\citep{roberts-cp}.

@@

We make use of and extend~\citet{MusicTools}, a library of small tools
for analysis and synthesis of music written in Agda, with the goal of
eventually being a dependently-typed alternative to Euterpea.
Previous work by \citet{cong-cp} using Music Tools expressed
first-species counterpoint with the Agda type system, so that an error
in counterpoint resulted in an Agda type error. Although this takes
advantage of Agda's native type checking, it has several
downsides. One is that the Agda errors may be difficult to interpret
as the corresponding musical error. Another is that it is difficult to
keep the types both simple and flexible.

>>>>

Edgar Var\`{e}se describes music as ``organized sound'', and
throughout history cultures have developed and applied systems of rules
and guidelines to govern the music they create, most notably the
Common Practice Period of Western music spanning the 17th to early
20th centuries. These rule systems are seldom absolute, and indeed
deliberate breaking of the rules is often part of the aesthetic, but they
roughly constrain the music they apply to and
give it a common form and sound.

@@

To describe rules of basic harmonic structure,
\citet{magalhaes-harmtrace} and their successors use dependent types,
for example to index chords by major or minor mode.  However Haskell
currently has limited support for dependent types, and requires many
extentions and tricks such as the use of singleton
types~\citep{eisenberg-singleton}.

In this paper we explore what can be done in the context of music by
using a programming language that offers full dependent types. We use
Agda~\citep{norell-phd} since it is fairly mature and
aims to be both a functional programming language and a proof
assistant. It also features a Haskell FFI so we can take advantage of
existing Haskell libraries, in partiular for MIDI generation.

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and 
formalize it in type theory. However this is difficult because most
``rules'' in music theory are not absolute, but rather suggestions of
varying degrees (themselves vaguely specified) of importance. Since
ultimately the resulting created music is more important than the
theory, an alternative is to attempt to indirectly deduce the rules
from music samples, using machine learning~\citep{huang-cp} or
some other statistical methods. Unfortunately the set of deduced
rules could sometimes be insufficient for generating good music.
As an example, \citet{roberts-cp} develop the Bach Doodle
harmonization tool using Bach chorals as training data, but still
observe high frequency of parallel 5th and 8th that are extremely
rare in Bach's music.

@@

We make use of and extend~\citet{MusicTools}, a library of small tools
for analysis and synthesis of music written in Agda, with the goal of
eventually being a dependently-typed alternative to Euterpea.
Previous work by \citet{cong-cp} using Music Tools expresses
first-species counterpoint with the Agda type system, so that an error
in counterpoint resulted in an Agda type error. Although this takes
advantage of Agda's native type checking, it has several
downsides. One is that the Agda errors may be difficult to interpret
as the corresponding musical error. Another is that it is difficult to
keep the types both simple and flexible.
